---
title: "STAT230 Final Project"
author: "Noah, Alex, Lucas"
date: "2024-03-04"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
# Loading in required libraries
library(mosaic)
library(ggformula)
library(car)
library(GGally)
library(broom)
library(ggeffects)
library(effects)
library(tidyr)
library(ggplot2)
library(rms)
```

```{r, loading data}
# Loading in data
by_player <- read.csv("games_details.csv")
games <- read.csv("games.csv")

# Converting player data in each game to be team data in each game
player_data <- by_player %>%
  group_by(GAME_ID, TEAM_ID) %>%
  summarise(FGM = sum(FGM, na.rm = TRUE),
            FGA = sum(FGA, na.rm = TRUE),
            FG3M = sum(FG3M, na.rm = TRUE),
            FG3A = sum(FG3A, na.rm = TRUE),
            FTM = sum(FTM, na.rm = TRUE),
            FTA = sum(FTA, na.rm = TRUE),
            OREB = sum(OREB, na.rm = TRUE),
            DREB = sum(DREB, na.rm = TRUE),
            total_reb = sum(REB, na.rm = TRUE),
            AST = sum(AST, na.rm = TRUE),
            STL = sum(STL, na.rm = TRUE),
            BLK = sum(BLK, na.rm = TRUE),
            TO = sum(TO, na.rm = TRUE),
            PF = sum(PF, na.rm = TRUE),
            PTS = sum(PTS, na.rm = TRUE),) %>%
  rename(team_id = TEAM_ID)

# Making an indicator variable for if the away team wins the game
games <- games %>% 
  mutate(AWAY_TEAM_WINS = 1 - HOME_TEAM_WINS)
```

```{r splitting data}
# Extracting variables pertinent for the home team
home_games <- games[,c(1,2,4,6,8,9,10,11,12,13,15,19,20,21)] %>%
  rename(PTS_ALLOWED = PTS_away,
         REB_ALLOWED = REB_away,
         AST_ALLOWED = AST_away
         )

# Extracting variables pertinent for the away team
away_games <- games[,c(1,2,5,6,8,12,13,15,16,17,18,19,20,22)] %>%
  rename(PTS_ALLOWED = PTS_home,
         REB_ALLOWED = REB_home,
         AST_ALLOWED = AST_home
         )
```

```{r mutating data}
# Making variable names uniform
home_games <- home_games %>%
  rename(points = PTS_home,
         fg_pct = FG_PCT_home,
         ft_pct = FT_PCT_home,
         fg3_pct = FG3_PCT_home,
         ast = AST_home,
         reb = REB_home,
         pts_allowed = PTS_ALLOWED,
         ast_allowed = AST_ALLOWED,
         reb_allowed = REB_ALLOWED,
         win = HOME_TEAM_WINS,
         team_id = HOME_TEAM_ID) %>%
  # Creating a home team variable
    mutate(home_team = 1)

# Making variable names uniform
away_games <- away_games %>%
  rename(points = PTS_away,
         fg_pct = FG_PCT_away,
         ft_pct = FT_PCT_away,
         fg3_pct = FG3_PCT_away,
         ast = AST_away,
         reb = REB_away,
         pts_allowed = PTS_ALLOWED,
         ast_allowed = AST_ALLOWED,
         reb_allowed = REB_ALLOWED,
         win = AWAY_TEAM_WINS,
         team_id = VISITOR_TEAM_ID) %>%
    # Creating an away team variable
  mutate(home_team = 0)
```

```{r, binding and completing data}
# Binding home and away data
games_data_complete <- rbind(home_games, away_games)

# Joining the additional data
final_data <- games_data_complete %>%
  left_join(player_data, by = c("GAME_ID", "team_id"))

# Creating a log(point allowed) variable for easier use
final_data <- final_data %>%
  mutate(pts_allowed_log = log(pts_allowed))
```

### **Exploratory Data Analysis**

```{r, warning = FALSE}
# Creating graphics of the number of attempted 3 point shots in each game
gf_boxplot(~FG3A, data=final_data) + 
  theme_bw()
gf_histogram(~FG3A, data=final_data) + 
  theme_bw()
```

```{r, warning = FALSE}
# Figure 1
gf_jitter(FG3A~SEASON, data=final_data) %>% gf_lm() + 
  labs(x="Season", y="3-point Field Goals Attempted", 
       title="Number of 3-Point field goals attempted in-game over different seasons") +      theme_bw()
```

```{r, warning = FALSE}
# Figure 1
gf_jitter(win~FG3A, data=final_data) %>% gf_lm() + 
  labs(x="3-point Field Goals Attempted", y="Win", 
       title="Win Vs 3FGA") +      theme_bw()
gf_jitter(win~reb, data=final_data) %>% gf_lm() + 
  labs(x="Rebounds", y="Win", 
       title="Win Vs Reb") +      theme_bw()

```

```{r}
# Additional plots exploring the data
gf_jitter(FG3A~pts_allowed, data=final_data) %>% gf_lm()+ theme_bw()
gf_jitter(fg3_pct~FG3A, data=final_data) %>% gf_lm()+ theme_bw() +
  labs(x="3-point Field Goals Attempted", y="3-Point Field Goal Percentage", title="3-point % against 3-point shots attempted")
gf_jitter(pts_allowed~SEASON, data=final_data) %>% gf_lm()+ theme_bw()
```

### **Fitting the Model, Wald's Tests, Deviance GoF test**

```{r}
# Fitting an inadequate attempt at a model
nba_glm <- glm(win ~ FG3A  + log(pts_allowed) + FTA  + OREB + DREB + TO + SEASON + FG3A:SEASON + log(pts_allowed):SEASON, data = final_data, family = "quasibinomial")
summary(nba_glm)
1 - pchisq(53624, 53094) #INADEQUACY
```

```{r}
# Fitting the model
nba_glm2 <- glm(win ~ FG3A + fg3_pct  + log(pts_allowed) + FTA  + reb + TO, data = final_data, family = "binomial")
summary(nba_glm2)
```

```{r}
# LRT Test
nba_glm2_reduced <-glm(win ~ log(pts_allowed) + FTA  + reb + TO, data = final_data, family = "binomial")
anova(nba_glm2_reduced,nba_glm2,test="LRT")
```

Table 1:

| Variable              | Coefficient | Standard Error | Wald's Test p-value |
|-----------------------|------------:|---------------:|--------------------:|
| Intercept             |      50.770 |          0.592 |            \< 0.001 |
| 3-point Attempts      |      0.0631 |        0.00151 |            \< 0.001 |
| 3-point Percentage    |      11.131 |          0.133 |            \< 0.001 |
| log(Points Allowed)   |     -13.271 |          0.137 |            \< 0.001 |
| Free Throws Attempted |       0.106 |        0.00172 |            \< 0.001 |
| Rebounds              |      0.0993 |        0.00201 |            \< 0.001 |
| Turnovers             |      -0.118 |        0.00306 |            \< 0.001 |

```{r}
# Goodness of Fit test for the second logistic model that indicates our model is adequate
1 - pchisq(45863, 53097)
```

### **Checking Model Assumptions**

```{r, message = FALSE, warning = FALSE}
# Checking for correlations between variables that could result in multicollinearity
final_data |> 
  ggpairs(columns = c("win", "FG3A", "fg3_pct", "pts_allowed", "FTA", "reb", "TO"))
```

```{r}
# Checking the Variance Inflation Factor
vif(nba_glm2)
# We find no evidence of significant multicollinearity with non-interaction terms.
```

```{r}
# Checking for outliers in our data

# Standardized residuals seem to be mostly between -2 and 2
plot(rstandard(nba_glm2), main="Standardized Residuals", ylab="Standardized Residuals", xlab="Index")

# Hat values are very small
plot(hatvalues(nba_glm2), main="Hat Values", ylab="Standardized Residuals", xlab="Index")

# Cook's distances all are very small as well
plot(cooks.distance(nba_glm2), main="Cooks D", ylab="Standardized Residuals", xlab="Index") 

# The 'influential' games are the games with game ids 10600015 and 20700065 which, when looking at the data, do not really seem to be outliers
cooks_d <- cooks.distance(nba_glm2)
influential_obs <- which(cooks_d > 0.0015)
print(influential_obs) 
```

```{r Binned Residual Plots}

# Augmenting Data
nba_res <- augment(nba_glm2, type.predict= "response") |>
  dplyr::mutate(.resp.resid = resid(nba_glm2, type= "response"))

# Avg. resids vs. predicted values
arm::binnedplot(nba_glm2$linear.predictors, nba_res$.resp.resid, 
                xlab = "Predicted odds", 
                main = "Binned Residual vs. Predicted Values", 
                col.int = FALSE)

# Individual Binned Residual Plots
arm::binnedplot(nba_res$FG3A, nba_res$.resp.resid, xlab = "FG3A", col.int = NULL)
#arm::binnedplot(nba_res$pts_allowed, nba_res$.resp.resid, xlab = "pts_allowed_log", col.int = NULL)
arm::binnedplot(nba_res$FTA, nba_res$.resp.resid, xlab = "FTA", col.int = NULL)
arm::binnedplot(nba_res$reb, nba_res$.resp.resid, xlab = "reb", col.int = NULL)
arm::binnedplot(nba_res$SEASON, nba_res$.resp.resid, xlab = "SEASON", col.int = NULL)
arm::binnedplot(nba_res$TO, nba_res$.resp.resid, xlab = "TO", col.int = NULL)
arm::binnedplot(nba_res$fg3_pct, nba_res$.resp.resid, xlab = "fg3_pct", col.int = NULL)
```

```{r}
# Checking model accuracy using concordant/discordant pairs
p_lrm <- lrm(win ~ FG3A  + log(pts_allowed) + FTA  + reb + TO, data = final_data) 
# SEASON IS NOT INCLUDED
# We appear to have a moderately strong association between our predictors and winning
p_lrm$stats[c("Dxy", "Gamma", "Tau-a")]
```

PLOTS: win v 3PA:

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("FG3A"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on 3 points shot", x="3FGA", y="Predicted probability of winning")
```

win v pts allowed:

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("pts_allowed", "SEASON"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on pts_allowed", x="pts_allowed", y="Predicted probability of winning")
```

win v free throws attempted

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("FG3A", "fg3_pct"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on 3FGA and Percentage", x="FG3A", y="Predicted probability of winning")

```

win v REB

```{r}
lm_preds <- ggpredict(nba_glm, terms = c("reb"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on reb", x="reb", y="Predicted probability of winning")
```

win v TO

```{r}
lm_preds <- ggpredict(nba_glm, terms = c("TO"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on TO", x="TO", y="Predicted probability of winning")
```

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("FG3A", "pts_allowed"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on 3 pointers attempted and points allowed", x="3-point field goals attempted", y="Predicted probability of winning")
```

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("fg3_pct", "pts_allowed"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on 3 pointers attempted and points allowed", x="3-point field goal percentage", y="Predicted probability of winning")
```

```{r}
lm_preds <- ggpredict(nba_glm2, terms = c("FG3A", "fg3_pct"))
plot(lm_preds, show_data = TRUE) + 
  labs(title="Predicted Probability of winning based on 3 pointers attempted and 3 pt %", x="3-point field goals attempted", y="Predicted probability of winning")
```

RESIDUAL PLOTS:

```{r}
nba_res <- augment(nba_glm, type.predict= "response") |>
  dplyr::mutate(.resp.resid = resid(nba_glm, type= "response"))

# Avg. resids vs. predicted values
arm::binnedplot(nba_glm$linear.predictors, nba_res$.resp.resid, 
                xlab = "Predicted odds", 
                main = "Binned Residual vs. Predicted Values", 
                col.int = FALSE)
```

```{r}
model_aug <- augment(nba_glm2) |>
  mutate(resid = resid(nba_glm2, type="response"))
arm::binnedplot(model_aug$FG3A, model_aug$resid, xlab="FG3A", col.int=NULL)
```

```{r}
p_lrm <- lrm(win ~  FG3A + fg3_pct  + log(pts_allowed) + FTA  + reb + TO, data = final_data) #SEASON IS NOT INCLUDED
p_lrm$stats[c("Dxy", "Gamma", "Tau-a")]
```
